---
title: "Nested Effects Models-based perturbation inference  \n
Inference of unobserved perturbations from gene expression profiles."
author: "Martin Pirkl, Niko Beerenwinkel"
date: "`r Sys.Date()`"
graphics: yes
header-includes:
    - \usepackage{xcolor}
output: BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{nempi}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

# Introduction

If many genes are perturbed in a population of cells, this can lead to
diseases like cancer. The perturbations can happen in different ways,
e.g. via mutations, copy number abberations or methylation. However,
not all perturbations are observed in all samples.

Nested Effects Model-based perturbation inference (NEM$\pi$) uses
observed perturbation profiles and gene expression data to infer(
unobserved perturbations and augment observed ones. The perturbed genes
(P-genes) are modelled in a signalling pathway $\phi$ and the genes with
observed gene expression (E-genes) are modelled with the attachment
$\theta$. If E-gene $j$ is attached to P-gene $i$, $j$ shows an effect
for a perturbation of P-gene $i$.

# Installation and loading
```{r global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, out.width="125%", fig.align="center",
                      strip.white=TRUE, warning=FALSE, tidy=TRUE,
                      #out.extra='style="display:block; margin:auto;"',
                      fig.height = 4, fig.width = 8, error=FALSE)
fig.cap0 <- "Heatmap of the simulated log odds. Effects are blue and no effects
are red. Rows denote the observed E-genes and columns the samples annoted by
P-genes. Each S-gene
has been perturbed in many cells. The E-genes are annotated as how they are
attached in the ground truth. E.g. E-genes named '1' are attached to S-gene
'1' in the ground truth."
fig.cap1 <- "Heatmap of the probabilsitic perturbation matrix."
paltmp <- palette()
paltmp[3] <- "blue"
paltmp[4] <- "brown"
palette(paltmp)
```

Use devtools to install the latest version from github or use the
BiocManahger to install the package from bioconductor.

```{r, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("cbg-ethz/nempi")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("nempi")
```
Load the package with the library function.
```{r}
library(nempi)
```

# Small example

We look at a small example for which we first simulate data and then
infer the unobserved perturbations. We draw a random network for the
perturbed genes (P-genes). Then we simulate gene expression for each
sample given the subset of P-genes that has been perturbed in the
sample. E.g., if P-gene A is upstream of P-gene B and P-gene A has
been perturbed, all E-genes attached to B also show an effect.

## Data simulation

```{r, fig.height=6, fig.width=10, fig.cap=fig.cap0}
library(mnem)
seed <- 42
Pgenes <- 10
Egenes <- 10
samples <- 100
uninform <- floor((Pgenes*Egenes)*0.1)
Nems <- mw <- 1
noise <- 1
multi <- c(0.2, 0.1)
set.seed(seed)    
simmini <- simData(Sgenes = Pgenes, Egenes = Egenes,
                  Nems = Nems, mw = mw, nCells = samples,
                  uninform = uninform, multi = multi,
                  badCells = floor(samples*0.1))
data <- simmini$data
ones <- which(data == 1)
zeros <- which(data == 0)
data[ones] <- rnorm(length(ones), 1, noise)
data[zeros] <- rnorm(length(zeros), -1, noise)
epiNEM::HeatmapOP(data, col = "RdBu", cexRow = 0.75, cexCol = 0.75,
                  bordercol = "transparent", xrot = 0,
                  dendrogram = "both")
```
The typical data input for NEM$\pi$ consists of a data matrix with
                  samples as columns and E-genes as rows. The
                  columns are either labeled by their perturbed gene(s)
                  or unlabeled (default: ""). After the data
                  simulation all samples are labeled. We unlabel $50\%$
                  of the sample and pretend we do not know, which P-gene
                  has been perturbed.

```{r}
lost <- sample(1:ncol(data), floor(ncol(data)*0.5))
colnames(data)[lost] <- ""
```

## Perturbation inference

We use NEM$\pi$ and other methods to infer the perturbations. We use
correlation and the area under the precision recal curve as the two
measures of accuracy.

```{r}
res <- nempi(data)
fit <- pifit(res, simmini, data)
print(c(Correlation = fit$cor, AUC = fit$auc))

ressvm <- classpi(data)
fit <- pifit(ressvm, simmini, data, propagate = FALSE)
print(c(Correlation = fit$cor, AUC = fit$auc))

resnn <- classpi(data, method = "nnet")
fit <- pifit(resnn, simmini, data, propagate = FALSE)
print(c(Correlation = fit$cor, AUC = fit$auc))

resrf <- classpi(data, method = "randomForest")
fit <- pifit(resrf, simmini, data, propagate = FALSE)
print(c(Correlation = fit$cor, AUC = fit$auc))
```

Comapred to support vector machines (svm), neural nets (nnet) and
random forest (rf) class prediction, NEM$\pi$ achieves a higher
accuracy.

## Prior matrix

Alternatively, we can also provide NEM$\pi$ with a probabilistic
perturbation matrix $\Gamma$ as a prior. In the rows are the potentially
perturbed
genes and in the columns are the samples. The entries are between $0$
and $1$, with the sum of all entries of a sample summing to $1$.

```{r, fig.height=6, fig.width=10, fig.cap=fig.cap1}
Gamma <- matrix(0, Pgenes, ncol(data))
rownames(Gamma) <- seq_len(Pgenes)
colnames(Gamma) <- colnames(data)
for (i in seq_len(Pgenes)) {
    Gamma[i, grep(paste0("^", i, "_|_", i,
                        "$|_", i, "_|^", i, "$"),
                 colnames(data))] <- 1
}
Gamma <- apply(Gamma, 2, function(x) return(x/sum(x)))
Gamma[is.na(Gamma)] <- 0

epiNEM::HeatmapOP(Gamma, col = "RdBu", cexRow = 0.75, cexCol = 0.75,
                  bordercol = "transparent", xrot = 0,
                  dendrogram = "both")

colnames(data) <- sample(seq_len(Pgenes), ncol(data), replace = TRUE)
res <- nempi(data, Gamma = Gamma)

fit <- nempi:::pifit(res, simmini, data)
print(c(Correlation = fit$cor, AUC = fit$auc))
```

# Session information

```{r}
sessionInfo()
```

# References:

Markowetz, F., Bloch, J., and Spang, R. (2005). Non-transcriptional
pathway features reconstructed from secondary effects
of rna interference. Bioinformatics, 21(21), 4026–4032.

Markowetz, F., Kostka, D., Troyanskaya, O. G., and Spang, R. (2007).
Nested effects models for high-dimensional phenotyping
screens. Bioinformatics, 23(13), i305–i312.

Pirkl, M., Beerenwinkel, N.; Single cell network analysis with a mixture
of Nested Effects Models, Bioinformatics, Volume 34, Issue 17, 1 September
2018,
Pages i964–i971, https://doi.org/10.1093/bioinformatics/bty602.

Ritchie ME, Phipson B, Wu D, Hu Y, Law CW, Shi W, Smyth GK (2015).
“limma powers differential expression analyses for RNA-sequencing and
microarray studies.” Nucleic Acids Research, 43(7), e47.
