---
title: "Nested Effects Models-based perturbation inference  \n
Inference of unobserved perturbations from gene expression profiles."
author: "Martin Pirkl, Niko Beerenwinkel"
date: "`r Sys.Date()`"
graphics: yes
output: BiocStyle::pdf_document
vignette: >
    %\VignetteIndexEntry{nempi}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

# Introduction

If many genes are perturbed in a population of cells, this can lead to
diseases like cancer. The perturbations can happen in different ways,
e.g. via mutations, copy number abberations or methylation. However,
not all perturbations are observed in all samples.

Nested Effects Model-based perturbation inference (NEM$\pi$) uses
observed perturbation profiles and gene expression data to infer
unobserv perturbations and augment observed ones.

# Installation and loading
```{r global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, out.width="125%", fig.align="center",
                      strip.white=TRUE, warning=FALSE, tidy=TRUE,
                      #out.extra='style="display:block; margin:auto;"',
                      fig.height = 4, fig.width = 8, error=FALSE)
fig.cap0 <- "Heatmap of the simulated log odds. Effects are blue and no effects
are red. Rows denote the observed E-genes and columns the S-genes. Each S-gene
has been perturbed in many cells. The E-genes are annotated as how they are
attached in the ground truth. E.g. E-genes named '1' are attached to S-gene
'1' in the ground truth."
paltmp <- palette()
paltmp[3] <- "blue"
paltmp[4] <- "brown"
palette(paltmp)
```
Install the package with the bioconductor manager package.
```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("nempi")
```
Load the package with the library function.
```{r}
library(nempi)
```

# Small example

We look at a small example for which we first simulate data and then
infere the unobserved perturbations.

## Data simulation

seed <- 27
Sgenes <- 3
Egenes <- 10
nCells <- 100
uninform <- 1
mw <- c(0.4, 0.3, 0.3)
Nems <- 3
noise <- 0.5
set.seed(seed)    
simmini <- simData(Sgenes = Sgenes, Egenes = Egenes,
                  Nems = Nems, mw = mw, nCells = nCells, uninform = uninform)
data <- simmini$data
ones <- which(data == 1)
zeros <- which(data == 0)
data[ones] <- rnorm(length(ones), 1, noise)
data[zeros] <- rnorm(length(zeros), -1, noise)
epiNEM::HeatmapOP(data, col = "RdBu", cexRow = 0.75, cexCol = 0.75,
                  bordercol = "transparent", xrot = 0, dendrogram = "both")
```{r, fig.height=6, fig.width=10, fig.cap=fig.cap0}

```

# Session information

```{r}
sessionInfo()
```

# References:

Markowetz, F., Bloch, J., and Spang, R. (2005). Non-transcriptional
pathway features reconstructed from secondary effects
of rna interference. Bioinformatics, 21(21), 4026–4032.

Markowetz, F., Kostka, D., Troyanskaya, O. G., and Spang, R. (2007).
Nested effects models for high-dimensional phenotyping
screens. Bioinformatics, 23(13), i305–i312.

Pirkl, M., Beerenwinkel, N.; Single cell network analysis with a mixture
of Nested Effects Models, Bioinformatics, Volume 34, Issue 17, 1 September 2018,
Pages i964–i971, https://doi.org/10.1093/bioinformatics/bty602.

Ritchie ME, Phipson B, Wu D, Hu Y, Law CW, Shi W, Smyth GK (2015).
“limma powers differential expression analyses for RNA-sequencing and
microarray studies.” Nucleic Acids Research, 43(7), e47.
